# Bootstrapping

```{r}
library("tidyverse")
library("broom")
library("modelr")
# devtools::install_github("jrnold/resamplr")
library("resamplr")
```


Non-parametric bootstrapping estimates standard errors and confidence intervals by resampling the observations in the data.

The `bootstrap` function in the [resamplr](https://github.com/resmaplr) implements simple non-parametric bootstrapping.[^bootstrap]
It generates `n` bootstrap replicates. 
```{r}
bsdata <- resamplr::bootstrap(car::Duncan, n = 1024)
glimpse(bsdata)
```
It returns a data frame with two columns an id, and list column, `sample` containing `r rdoc("modelr", "resample")` objects.
The `resample` objects consist of two elements: `data`, the data frame; `idx`, the indexes of the data in the sample.
```{r}
bsdata[["sample"]][[1]]
```
Since the `data` object hasn't changed it doesn't take up any additional memory until subsets are created, allowing for the creation of `lazy` subsamples of a dataset.
A `resample` object can be turned into a data frame with `as.data.frame`:
```{r}
as.data.frame(bsdata[["sample"]][[1]])
```

[^bootstrap]: This is based on the function `r rdoc("modelr", "bootstrap")` in the `r rpkg("modelr")` package, but includes more options. The **broom** package also provides a `bootstrap` function.

To generate standard errors for a statistic, estimate it on each bootstrap replicate.

Suppose, we'd like to calculate robust standard errors for the regression coefficients in this regresion:
```{r}
mod <- lm(prestige ~ type + income + education, data = car::Duncan)
mod
```

Since we are interested in the coefficients, we need to re-run the regression with `lm`, extract the coefficients to a data frame using `tidy`, and return it all as a large data frame.
For one bootstrap replicate this looks like,
```{r}
lm(prestige ~ type + income + education, data = as.data.frame(bsdata$sample[[1]])) %>%
  tidy() %>%
  select(term, estimate)
```
Note that the coefficients on this regression are slightly different than those in the original regression.

```{r}
bs_coef <- map_df(bsdata$sample, function(dat) {
  lm(prestige ~ type + income + education, data = dat) %>%
    tidy() %>%
    select(term, estimate)
})
```

There are multiple methods to estimate standard errors and confidence intervals using the bootstrap replicate estimates.
Two simple ones are are

1. Use the standard deviation of the boostrap estimates as $\hat{se}(\hat{\beta})$ instead of those produces by OLS. The confidence intervals are generated using the OLS coefficient estimate and the bootstrap standard errors, $\hat{\beta}_{OLS} \pm t_{df,\alpha/2}^* \hat{se}_{boot}(\hat{\beta})$
2. Use the quantiles of the bootstrap estimates as the endpoints of the confidence interval. E.g. the 95% confidence interval uses the 2.5th and 97.5th quantiles of the bootstrap estimates.

The first (standard error) method requires less bootstrap replicates. The quantile method allows for
asymmetric confidence intervals, but is noisier (the 5th and 95th quantiles vary more by samples) and requires more bootstrap replicates to get an accurate estimate. 

The bootstrap standard error confidence intervals:
```{r}
alpha <- 0.95
tstar <- qt(1 - (1 - alpha / 2), df = mod$df.residual)
bs_est_ci1 <- 
  bs_coef %>%
  group_by(term) %>%
  summarise(std.error = sd(estimate)) %>%
  left_join(select(tidy(mod),
                   term, estimate,
                   std.error_ols = std.error),
            by = "term") %>%
  mutate(
   conf.low = estimate - tstar * std.error,
   conf.high = estimate + tstar * std.error    
  )
select(bs_est_ci1, term, conf.low, estimate, conf.high)
select(bs_est_ci1, term, std.error, std.error_ols)
```
The quantile confidence intervals:
```{r}
alpha <- 0.95
bs_coef %>%
  group_by(term) %>%
  summarise(
   conf.low = quantile(estimate, (1 - alpha) / 2),
   conf.high = quantile(estimate, 1 - (1 - alpha) / 2)
  ) %>%
  left_join(select(tidy(mod),
                   term, estimate),
            by = "term") %>%
  select(term, estimate)
```


See the **boot** package (and other cites TODO) for more sophisticated methods of generating standard errors and quantiles.

The package [resamplr](https://github.com/jrnold/resamplr) includes more methods using `resampler` objects.
The package `r rpkg("boot")` implements many more bootstrap methods [@Canty2002a].
