# Problems with Errors


## Prerequisites

In addition to tidyverse pacakges, this chaper uses the `r rpkg("sandwich")` and `r rpkg("lmtest")` packages which provide robust standard errors and tests that use robust standard errors.

```{r message=FALSE}
library("sandwich")
library("lmtest")
library("tidyverse")
library("broom")
library("modelr")
```


## Heteroskedasticity

$$
\hat{\beta} = (\mat{X}\T \mat{X})^{-1} \mat{X}\T \vec{y}
$$
and 
$$
\Var(\vec{\epsilon}) = \mat{\Sigma}
$$
is the variance-covariance matrix of the errors.

Assumptions 1-4 give the expression for the sampling variance,
$$
\Var(\hat{\beta}) = (\mat{X}'\mat{X})^{-1} \mat{X}\T \mat{\Sigma} \mat{X} (\mat{X}\T \mat{X})^{-1}
$$
under homoskedasticity, 
$$
\mat{\Sigma} = \sigma^2 \mat{I},
$$
so the the variance-covariance matrix simplifies to
$$
\Var(\hat{\beta} | X) = \sigma^2 (\mat{X}\T \mat{X})^{-1}
$$

Homoskedastic:
$$
\Var(\vec{\epsilon} | \mat{X}) = \sigma^2 I = 
\begin{bmatrix}
\sigma^2 & 0        & 0      & \cdots & 0 \\
0        & \sigma^2 & 0      & \cdots & 0 \\
\vdots   & \vdots   & \vdots & \ddots & \vdots \\
\sigma^2 & 0        & 0      & \cdots & \sigma^2 
\end{bmatrix}
$$

Heteroskedastic
$$
\Var(\vec{\epsilon} | \mat{X}) = \sigma^2 I = 
\begin{bmatrix}
\sigma_1^2 & 0        & 0      & \cdots & 0 \\
0        & \sigma_2^2 & 0      & \cdots & 0 \\
\vdots   & \vdots   & \vdots & \ddots & \vdots \\
\sigma^2 & 0        & 0      & \cdots & \sigma_n^2 
\end{bmatrix}
$$
- independent, since the only non-zero values are on the diagonal, meaning that there are no correlated errors between observations
- non-identical, since the values on the diagonal are not equal, e.g. $\sigma_1^2 \neq \sigma_2^2$.
- $\Cov(\epsilon_i, \epsilon_j | \mat{X}) = 0$
- $\Var(\epsilon_i | \vec{x}_i) = \sigma^2_i$

```{r}
tibble(
  x = runif(100, 0, 3),
  `Homoskedastic` = rnorm(length(x), mean = 0, sd = 1),
  `Heteroskedasticity` = rnorm(length(x), mean = 0, sd = x)
) %>%
  gather(type, `error`, -x) %>%
  ggplot(aes(x = x, y = error)) +
  geom_hline(yintercept = 0, colour = "white", size = 2) +
  geom_point() +
  facet_wrap(~ type, nrow = 1)

```

Consequences

- $\hat{\vec{\beta}}$ are still unbiased and consistent estimators of $\vec{\beta}$
- Standard error estimates are **biased**, likely downward, meaning that the estimated standard errors will be smaller than the true standard errors (too optimistic).
- Test statstics won't be distributed $t$ or $F$
- $\alpha$-level tests will have Type I errors $\neq \alpha$
- Coverage of confidence intervals will not be correct.
- OLS is not BLUE

Visual diagnostics

- Plot residuals vs. fitted values
- Spread-location plot.
  - y: square root of absolute value of residuals
  - x: fitted values
  - loess trend curve
  
Dealing with NCV

- Transform the dependent variable
- Model the heteroskedasticity using WLS
- Use an estimator of $\Var(\hat{\beta} | \mat{X})$ that is **robust** to heteroskedasticity
- Admit we are using the **wrong model** and use a different model


The standard way to "fix" robust heteroskedasticity is to use so-called "robust" standard errors, more formally called Heteroskedasticity Consistent (HC), and heteroskedasticity and Autocorrelation Consistent standard errors.
HC and HAC errors are implemented in the R package `r rpkg("sandwich")`.
See @Zeileis2006a and Zeileis2004a for succint discussion of the estimators themselves and examples of their usage.

With robust standard errors, the coefficients of the model are estimated using `lm()`. 
Then a HC or HAC variance-covariance matrix is computed which corrects for heteroskedasticity (and autocorrelation).


### Example: Duncan's Occupation Data


```{r}
mod <- lm(prestige ~ income + education + type, data = car::Duncan)
```
The classic OLS variance covariance matrix is,
```{r}
vcov(mod)
```
and the standard errors are the diagonal of this matrix
```{r}
sqrt(diag(vcov(mod)))
```

Now, use `vcovHC` to estimate the "robust" variance covariance matrix
```{r}
vcovHC(mod)
```
and the robust standard errors are the diagonal of the matrix
```{r}
sqrt(diag(vcovHC(mod)))
```
Note that the robust standard errors are **larger** than the classic standard errors; this is almost always the case.

If you need to use the robust standard errors to calculate t-statistics or p-values.
```{r}
coeftest(mod, vcovHC(mod))
```


**TODO** An example that uses `vcovHAC()` to calculate heteroskedasticity and autocorrelation consistent standard errors.

#### WLS vs. White's esimator

WLS:

- different estimator for $\beta$: $\hat{\beta}_{WLS} \neq \hat{\beta}_{OLS}$
- With known weights:
    - efficient
    - $\hat{\se}(\hat{\beta}_{WLS})$ are consistent
- If weights aren't known ... then biased for both $\hat{\beta}$ and standard errors.

White's esimator (heteroskedasticity consistent standard errors):

- uses OLS estimator for $\beta$
- consistent for $\Var(\hat{\beta})$ for any form of heteroskedasticity
- relies on consistency and large samples, and for small samples the performance may be poor.

### Notes

An additional use of robust standard errors is to diagnose potential model fit problems.
The OLS line is still the minimum squared error of the population regression, but large differences may suggest that it is a poor approximation.
@KingRoberts2015a suggest a formal test for this using the variance-covariance matrix.

- Note that there are other functions that have options to input variance-covariance matrices along with the `lm` object in order to use robust standard errors with that test or routine.
- Heteroskedastic consistent standard errors can be used with MLE models [@White1982a]. However, this is 
- More generally, robust standard errors can be controversial: @KingRoberts2015a suggest using them to diagnose model fit problems.


## Non-normal Errors

This really isn't an issue. Normal errors only affect the standard errors, and only if the sample size is small. Once there is a reasonably large residual degrees of freedom (observations minus parameters), the CLT kicks in and it doesn't matter. 

If you are concerned about non-normal error it may be worth asking:

- Is the functional form, especially the form of the outcome varaible, correct?
- Is the conditional expected value ($Y | X$) really the best estimand? That's what the regression is giving you, but the conditional median or other quantile may be more appropriate for your purposes.

To diagnose use a qq-plot of the residuals against a normal distribution.

## Clustered Standard Errors

Clustering is when observations within groups are correlated.

See @CameronMiller2015a for complete review, and @EsareyMenger2017a for a recent discussion of clustered standard errors in political science.

Suppose there are $J$ equal sized clusters with $m$ units from each cluster, and total sample size of $J m$.
The mean of a vector $y$ is $\hat{y}$, and its standard error is [@GelmanHill2007a, p. 447]
$$
\se(\bar{y}) = \sqrt{\sigma^2_y / n + \sigma^2_{\alpha}
/ J},
$$
where $\sigma^2_{\alpha}$ is the variance of the cluster level means, and $\sigma^2_{y}$ is variance of the intra-cluster residuals.
This can also be rewritten as,
$$
\se(\bar{y}) = \sqrt{\sigma^2_{total} / J(1 + (m - 1)) ICC},
$$
where $\sigma^2_{total} = \sigma^2_{\alpha} + \simga^2_y$, and the $ICC$ is the intra-class correlation, which is the fraction of total variance accounted for by the between group variation,
$$
ICC = \frac{\sigma_{\alpha}^2}{\sigma^2_{\alpha} + \sigma^2_{y}} .
$$

How does the standard error of $\bar{y}$ change with the value of ICC? When ICC is 0? When ICC is 1? 


### Software

The `r rpkg("sandwich")` package for HC and HAC standard errors does not include clustered standard errors.

- `r rpkg("plm")` includes wraps the **sandwich** implmementation of robust standard errors for panels (clustered).
- `r rpkg("clubSandwich")` implements several newer estimators of clustered standard errors, especially for small samples.
- `r rpkg("clusterSE")` implements multiple methods appropriate for small numbers of clusters.

### References

- @CameronGelbachMiller2011a provide a method for two- and multi-way clustering
- @CameronGelbachMiller2008a introduces bootstrap methods for 
clustered standard errors.
