---
title: "Untitled"
author: "Jeffrey Arnold"
date: "5/8/2017"
output: html_document
---

$$
\DeclareMathOperator{diag}{diag}
$$

# Prerequisites

Before starting, load the necessary packages and define some functions that will be used later.

```{r message=FALSE}
library("tidyverse")
library("stringr")
library("broom")
library("modelr")
```

Combine standard deviation and correlation to make a covariance function.
```{r}
sdcor2cov <- function(s, r = diag(length(s))) {
  s <- diag(s, nrow = length(s), ncol = length(s))
  s %*% r %*% s
}
```
The $n \times n$ covariance matrix $\Sigma$ can be decomposed into the $n \times 1$ standard deviation vector $\sigma$ and correlation matrix $R$,
$$
\Sigma = \diag(\sigma) R \diag(\sigma)
$$

Generate a data frame with possibly correlated, multivariate normal variables.
The multivariate normal distribution is specified in terms of a mean (`mu`), 
standard deviation for variable (`sigma`), and the correlation matrix between variables,
(`R`):
```{r}
mvnorm_df <- function(n, mu, sigma = rep(1, length(mu)), R = diag(length(sd)),
                      empirical = TRUE) {
  as_tibble(MASS::mvrnorm(n, mu = mu, Sigma = sdcor2cov(sigma, R), empirical = empirical))
}
```

Find a regression standard error that will produce a given $R^2$.
For a normal linear model, given a sample $\hat{y}$ find the regression standard error, $\sigma^2$ to produce the desired $R^2$.
```{r }
r2_to_sigma <- function(y, r2) {
  # Var(Y) = E(Var(Y|X)) + Var(E(Y|X))
  ssm <- sum((y - mean(y)) ^ 2)
  sse <- (1 - r2) / r2 * ssm
  # return sigma (assume population n)
  sqrt(sse / n)
}
```


Consider four variables; a dependent variable $Y$, an endogenous variable $X$, an exogenous variable $Z$, and an instrumental variable $W$.
Suppose that the relationship is,
$$
Y = \alpha + \beta X + \gamma Z + \epsilon
$$
The correlations between the variables are such that $X$ is not correlated with the error $\epsilon$, conditional on $Z$, and 
the instrumental variable $W$ is correlated with $X$, but not correlated with the error $\epsilon$ after controlling for $Z$ (only correlated with $Y$ through the endogenous variable $X$):
$$
\begin{aligned}[t]
\Cov(X, \epsilon | Z) = 0 \\
\Cov(X, W) \neq 0 \\
\Cov(Y, \epsilon | Z) = 0
\end{aligned}
$$

Let's consider binary endogenous and exogenous variables




