# Many Models

These notes build off of the topics discussed in the chapter [Many Models](http://r4ds.had.co.nz/many-models.html) in *R for Data Science*.
It uses functionals (`map()` function) for iteration, string functions,
and list columns in data frames.

## Prerequisites

```{r}
library("tidyverse")
library("stringr")
library("broom")
```


## Programming with Formulas


In these examples, we'll use the `r rdoc("car", "Prestige")` dataset in the 
`r rpkg("car")` package.
```{r}
Prestige <- car::Prestige
```
Each observation is an occupation, and contains the prestige score of the 
occupation from a survey, and the average education, income, percentage of women, and type of occumpation. 
```{r}
glimpse(Prestige)
```
We will run several regressions with prestige as the outcome variable,
and the over variables are explanatory variables.

## Programming with Formulas

In R, the formulas are objects (of class `"formula"`).
That means we can program on them, and importantly, perhaps avoid excessive
copying and pasting if we run multiple models.

A formula object is created with the `~` operator:
```{r}
f <- prestige ~ type + education
class(f)
f
```

A useful function for working with formulas is [update](`r rdoc_url("stats", "update.formula")`).
The `update` function allows you to easiy
```{r}
# the . is replaced by the original formula values
update(f, . ~ income)
update(f, income ~ .)
update(f, . ~ . + type + women)
```
Also note that many types of models have `update` method which will rerun the model with a new formula.
Sometimes this can help computational time if the model is able to reuse some previous results or data.

You can also create formulae from a character vector
```{r}
as.formula("prestige ~ income + education")
```

This means that you can create model formulae programmatically
which is useful if you are running many models, or simply to keep
the logic of your code clear.
```{r}
xvars <- c("type", "income", "education")
as.formula(str_c("prestige", "~", str_c(xvars, collapse = " + ")))
```

Often you will need to run multiple models.
Since most often the only thing that changes between models is the
formula (the outcome or response variables), storing the formula
in a list, and then running the models by iterating through the list
is a clean strategy for estimating your models.
```{r}
xvar_list <- list(c("type"),
                  c("income"),
                  c("education"),
                  c("type", "income"),
                  c("type", "income", "education"))
formulae <- vector("list", length(xvar_list))
for (i in seq_along(xvar_list)) {
  formulae[[i]] <- str_c("prestige ~ ",
                         str_c(xvar_list[[i]], collapse = " + "))
}
formulae
```

Alternatively, create this list of formulae with a functional,
```{r}
make_mod_f <- function(x) {
  str_c("prestige ~ ", str_c(x, collapse = " + "))  
}
formulae <- map(xvar_list, make_mod_f)
```

Now that we have the various model formulae we want to run, we can 

Run a single model that returns a data frame with a single row and column:

- `mod`: a list column with `lm` object with the fitted model. I set `model = FALSE`
    because by default an `lm` model stores the data used to estimte it.
    This is convenient, but if you are estimating many models, it can start
    taking up space.

```{r}
run_reg <- function(f) {
  mod <- lm(f, data = Prestige, model = FALSE)
  data_frame(mod = list(mod))
}

ret <- run_reg(formulae[[1]])
ret[["mod"]][[1]]
```

It doesn't make much sense to store that as a data frame on its own, but with
multiple inputs it will be useful.

Now, run `run_reg` for each formula in `formulae` using `map_df` to return 
the results as a data frame with a list column, `mod`, containing the `lm` objects.
```{r}
prestige_fits <- map_df(formulae, run_reg, .id = ".id")
prestige_fits
```

From here, it is easy to extract parts of the models that.
To extract the original formulas and add them to the data set,
run `formula()` on each `lm` object using `map`, and then convert
it to a character string using `deparse`:
```{r}
prestige_fits <- prestige_fits %>%
  mutate(formula = map_chr(mod, ~ deparse(formula(.x))))
prestige_fits$formula
```

Get a data frame of the coefficients for all models using `tidy` and unnest:
```{r}
mutate(prestige_fits, x = map(mod, tidy)) %>% 
  unnest(x)
```

Get a data frame of model summary statistics for all models using `glance`,
```{r}
mutate(prestige_fits, x = map(mod, glance)) %>%
  unnest(x)
```

