# Miscellaneous Regression Stuff

```{r}
library("tidyverse")
library("stringr")
```


## Anscombe quartet

```{r}
anscombe_tidy <-
  anscombe %>%
  mutate(obs = row_number()) %>%
  gather(variable_dataset, value, -obs) %>%
  separate(variable_dataset, c("variable", "dataset"), sep = c(1)) %>%
  spread(variable, value) %>%
  arrange(dataset, obs)

```

What are summary statistics of the four Anscombe datasets?
```{r}
ggplot(anscombe_tidy, aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ dataset, ncol = 2)
```

What are the mean, standard deviation, correlation coefficient, and regression
coefficients of each line? 
```{r}
anscombe_summ <-
  anscombe_tidy %>%
  group_by(dataset) %>%
  summarise(
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x = sd(x),
    sd_y = sd(y),
    cov = cov(x, y),
    cor = cor(x, y),
    coefs = list(coef(lm(y ~ x, data = .)))
  ) %>%
  mutate(
    intercept = map_dbl(coefs, "(Intercept)"),
    slope = map_dbl(coefs, "x")
  ) %>%
  select(-coefs)

```

WUT? They are the same? But they look so different. Of course that was the point ...

Since this all revolves around covariance, lets calculate the values of $x_i - \bar{x}$,
$y_i - \bar{y}$, and $(x_i - \bar{x}) (y_i - \bar{y})$ for each obs for each variable.
```{r}
anscombe_tidy <-
  anscombe_tidy %>%
  group_by(dataset) %>%
  mutate(mean_x = mean(x),
         diff_mean_x = x - mean_x,
         mean_y = mean(y),
         diff_mean_y = y - mean_y,
         diff_mean_xy = diff_mean_x * diff_mean_y,
         quadrant = 
           if_else(
             diff_mean_x > 0, 
             if_else(diff_mean_y > 0, 1, 2),
             if_else(diff_mean_y > 0, 4, 3),
           ))
```


```{r}
ggplot(anscombe_tidy, aes(x = x, y = y,
                          size = abs(diff_mean_xy),
                          colour = factor(sign(diff_mean_xy)))) +
  geom_point() +
  geom_hline(data = anscombe_summ,
             aes(yintercept = mean_y)) +
  geom_vline(data = anscombe_summ,
             aes(xintercept = mean_x)) +  
  facet_wrap(~ dataset, ncol = 2)
```

```{r}
ggplot(anscombe_tidy, aes(x = x, y = y, colour = factor(sign(diff_mean_xy)))) +
  geom_point() +
  geom_segment(mapping = aes(xend = mean_x, yend = mean_y)) +
  geom_hline(data = anscombe_summ,
             aes(yintercept = mean_y)) +
  geom_vline(data = anscombe_summ,
             aes(xintercept = mean_x)) + 
  facet_wrap(~ dataset, ncol = 2)
```

```{r}
ggplot(anscombe_tidy, aes(x = x, y = y, colour = factor(sign(diff_mean_xy)))) +
  geom_point() +
  geom_segment(mapping = aes(xend = x, yend = mean_y)) +
  geom_segment(mapping = aes(xend = mean_x, yend = y)) +  
  geom_hline(data = anscombe_summ,
             aes(yintercept = mean_y)) +
  geom_vline(data = anscombe_summ,
             aes(xintercept = mean_x)) +  
  facet_wrap(~ dataset, ncol = 2)

```


## Correlation Plots

```{r echo=FALSE}
# TODO: translate the original R code to produce a dataset and 
# reproduce plot in ggplot2
library("mvtnorm")
mv_normal <- function(cor = 0.8, n = 1000) {
    out <- as_tibble(rmvnorm(n, c(0, 0), matrix(c(1, cor, cor, 1), ncol = 2)))
    colnames(out) <- c("x", "y")
    tibble(title = str_c("Bivariate Normal (rho = ", cor, ")"),
           cor = round(cor(out$x, out$y), 1),
           data = list(out))
}

rotate <- function(t, X) {
  out <- as_tibble((X %*% matrix(c(cos(t), sin(t), -sin(t), cos(t)), ncol = 2)))
  colnames(out) <- c("x", "y")
  tibble(title = str_c("Rotation (", t, ")"),
         cor = round(cor(out$x, out$y), 1),
         data = list(out))
}

line_rotation <- function(t, n = 1000) {
  X <- rmvnorm(n, c(0, 0), matrix(c(1, 1, 1, 1), ncol = 2))
  out <- as_tibble((X %*% matrix(c(cos(t), sin(t), -sin(t), cos(t)), ncol = 2)))
  colnames(out) <- c("x", "y")
  tibble(title = str_c("Rotation (", t, ")"),
         cor = if (t == pi / 4) NA else round(cor(out$x, out$y), 1),
         data = list(out))  
}

wave <- function(n = 1000) {
  x <- runif(n, -1, 1)
  y <- 4 * (x ^ 2 - 1 / 2) ^ 2 + runif(n, -1, 1) / 3
  tibble(title = "y = 4 * (x ^ 2 - 1) / 2 ^ 2 + U(-1/3, 1/3)",
         cor = round(cor(x, y), 1),
         data = list(tibble(x = x, y = y)))
}

crescent <- function(n = 1000) {
  x <- runif(n, -1, 1)
  y <- 2 * x ^ 2 + runif(n, -1, 1)
  tibble(title = "y = 2 * x ^ 2 + U(-1, 1)",
         cor = round(cor(x, y), 1),
         data = list(tibble(x = x, y = y)))
}

double_crescent <- function(n = 1000) {
  x <- runif(n, -1, 1)
  y <- (x^2 + runif(n, 0, 1/2)) * sample(seq(-1, 1, 2), n, replace = TRUE)
  tibble(
    title = "y = c(-1, 1) * (x^2 + U(0, 1/2))",
    cor = round(cor(x, y), 1),
    data = list(tibble(x = x, y = y))
  )
}

circle <- function(n = 1000) {
  x <- runif(n, -1, 1)
  y = cos(x * pi) + rnorm(n, 0, 1 / 8)
  x = sin(x * pi) + rnorm(n, 0, 1 / 8)
  tibble(
    title = "x = cos(x * pi), y = cos(x * pi)",
    cor = round(cor(x, y), 1),
    data = list(tibble(x = x, y = y))
  )
}

four_mvnormals <- function(x, y = x, n = 1000) {
   out <- rbind(rmvnorm(n / 4, c( x,  y)),
                rmvnorm(n / 4, c(-x,  y)),
                rmvnorm(n / 4, c(-x, -y)),
                rmvnorm(n / 4, c( x, -y))) %>%
     as_tibble()
   colnames(out) <- c("x", "y")
   tibble(title = "Four multivariate normals",
          cor = round(cor(out$x, out$y), 1),
          data = list(out))
}

corr_examples <- bind_rows(
  purrr::map_df(c(1, 0.8, 0.4, 0.0, -0.4, -0.8, -1.0), mv_normal),
  purrr::map_df(c(0, pi / 12, pi / 6, pi / 4, pi / 2 - pi / 6, pi / 2 - pi / 12, pi / 2),
              line_rotation),
  wave(),
  rotate(-pi / 8, cbind(runif(1000, -1, 1), runif(1000, -1, 1))),
  rotate(-pi / 4, cbind(runif(1000, -1, 1), runif(1000, -1, 1))),
  crescent(),
  double_crescent(),
  circle(),
  four_mvnormals(3)
) %>%
  mutate(.id = row_number())

corr_examples %>%
  unnest(data) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_text(data = corr_examples,
            mapping = aes(label = sprintf("%.1f", cor), x = Inf, y = Inf),
            vjust = "inward", hjust = "inward") +
  facet_wrap(~ .id, ncol = 7)

```

<a title="By DenisBoigelot, original uploader was Imagecreator (Own work, original uploader was Imagecreator) [CC0], via Wikimedia Commons" href="https://commons.wikimedia.org/wiki/File%3ACorrelation_examples2.svg"><img width="2000" alt="Correlation examples2" src="https://upload.wikimedia.org/wikipedia/commons/d/d4/Correlation_examples2.svg"/></a>
