# Omitted Variable Bias

## Prerequisites

This chapter uses the `r rdoc("car", "Duncan")` data set in the `r rpkg("car")` package.
```{r message=FALSE}
library("car")
library("tidyverse")
```

## Simpson's Paradox

Before considering the more general phenomena of omitted variable bias, we'll discuss 
[Simpson's Paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox).[^simpson1]
This when a trend or relationship appears when data is disaggregated into groups, but that trend or relationship either disappears or reverses when the data are aggregated.

A famous real-world case of this @BickelHammelOConnell1975a, which analyzes a claim of sex bias in graduate admissions against UC-Berkeley in the 1970s.
In 1973, 8,442 men and 4,321 women applied for admission to Berkeley graduate programs.
In aggregate, UC Berkeley admitted 44% of men and 35% of women applicants, seemingly supporting that claim.
However, when the admissions rates were disaggregated by graduate department, the acceptance rates by department were not, on average, different.
What is going on?
On average, more women applied to more selective (higher rejection rate) departments than men. 

The dataset `r rdoc("datasets", "UCBAdmissions")` in the `r rpkg("datasets")` package
contains data for the largest 6 programs.
```{r}
data("UCBAdmissions", package = "datasets")
admissions <- as_tibble(UCBAdmissions) %>%
  spread(Admit, n) %>%
  mutate(applicants = Admitted + Rejected,
         accepted = Admitted / applicants)

ggplot(admissions, aes(x = Dept, y = accepted, size = applicants, colour = Gender)) +
  geom_point()
```
```{r}
select(admissions, Dept, Gender, applicants, accepted) %>%
  arrange(Dept, Gender)
```

An interactive visualization is this UC Berkeley VUD Lab visualization:  [Simpson's Paradox](http://vudlab.com/simpsons/).
**Stop. Go to that link. Explore that visualization, and build your intution**


[^simpson1]: However, omitted variable bias only makes sense as a concept when the
  regression has a causal or structural interpretation. Simpson's paradox is
  an aggregation affect that can occur with conditioning even when there is 
  no causal interpretation of the model.
  
For some other examples see @Moore2005a, @Wagner1982a, [Wikipedia](https://en.wikipedia.org/wiki/Simpson%27s_paradox), @JuliousMullee1994a (Kidney stone treatment).


## Omitted Variable Bias

Suppose that the population model is,
$$
Y_i = \beta_0 + \beta_1 X_i + \beta_2 Z_i + \epsilon_i ,
$$
but given a sample, we run a regression with only $\vec{x}$ and not $\vec{z}$.
$$
y_i = \hat{\beta}_0 + \hat{\beta}_1 x_i + \hat{\epsilon}_i .
$$

What is the relationship between $\beta_1$ and $\hat{\beta}_1$?
Is $\hat{\beta}_1$ an unbiased estimator of $\beta_1$ ? 



$$
\text{omitted variable bias} =  (\text{effect of $Z_i$ on $Y_i$}) \times (\text{effect of $X_i$ on $Z_i$})
$$


What does the omitted variable bias 
An *irrelevant variable* is one that is uncorrelated with $Y_i$, meaning that its 
population coefficient is 0.
Suppose $Z_i$ is an irrelevant variable,
$$
Y_i = \beta_0 + \beta_1 X_i + 0 \times Z_i = \epsilon_i
$$

In this case OLS is unbiased ...
$$
\begin{aligned}[t]
\E(\hat\beta_0) &= \beta_0 \\
\E(\hat\beta_1) &= \beta_1 \\
\E(\hat\beta_2) &= 0
\end{aligned}
$$

However, including an irrelevant variable will increase the standard errors for $\hat{\beta}_1$. Why?


Consider the linear regression model,
$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i.
$$
What if we included $X_i$ twice?
$$
Y_i = \tilde\beta_0 + \tilde\beta_1 X_i + \tilde\beta_2 X_i + \epsilon_i.
$$
Clearly, any combination of $\tilde\beta_1$ and $\tilde\beta_2$ where
$$
\tilde\beta_1 + \tilde\beta_2 =\beta_1 
$$
will fit the model as well as any other.

Consider cases of 

- bivariate OLS with "effective" number of observations
- continuous OLS


----------------- ------------------- ------------------ ------------------
$\Cov(X_1, X_2)$  $\Cov(X_2, Y) > 0$  $\Cov(X_2, Y) = 0$ $\Cov(X_2, Y) < 0$
----------------- ------------------- ------------------ ------------------
$> 0$             +                   0                  -
$0$               0                   0                  0
$< 0$             -                   0                  +
----------------- ------------------- ------------------ ------------------

In practice, this is the primary problem of many papers and papers. That is because it biases the coefficient of interest. Reviewers and discussants will often ask about whether you have considered "controlling" for *insert variable here*.

Although these may be legitimate concerns, not all reviewerss understand the purpose of controls variables so some of these may not be legitimate, and in fact harmful.
There two arguments to consider when controlling for a variable.

1. The omitted variable has to plausibly be correlated with *both* the variable of interest *and* the outcome variable, and the burden is on the reviewers to provide at a confounding variable and plausible relationships. Simply stating that there *could* be an unobservable variable is trivially true, uninteresting, and not a fatal critique. That said, the plausibility of a causal claim would be higher if with methods less susceptible to unobserved confounders, such as experiments, instrumental variables, regression discontinuity, and difference-in-differences.

2. The omitted variable should be not be "post treatment" variable. If the omitted variable should not be one of the causal pathways by which $X$ affects $Y$, it should not be controlled for. If $Z$ affects the values of $X$ and also affects $Y$, then it needs to be controlled for.

How to assess the potential magnitude of omitted variable bias?

1. **Informal method**. This is the methods that you see in many empirical papers. They estimate the model including different control variables. The less sensitive the coefficient(s) of the variables of interest are to the inclusion of control variables, the more plausible it is that the variable of interest also not sensitive to unobserved confounders [@AngristPischke2014a]. @Oster2016a states

    > A common heuristic for evaluating the robustness of a result to omitted variable bias concerns is to look
    > at the sensitivity of the treatment effect to inclusion of observed controls. In three top general interest
    > economics journals in 2012, 75% of non-experimental empirical papers included such sensitivity analysis. The
    > intuitive appeal of this approach lies in the idea that the bias arising from the observed controls is informative
    > about the bias that arises from the unobserved ones.

    Note that what is important is that the magnitude of the *coefficient* is stable to the inclusion of controls, not that the coefficient remains statistically significant.

2. **Formal methods:**  @BellowsMiguel2009a propose the following simple statistic to assess the magnitude of omitted variable bias:
$$
\delta = \frac{\hat{\beta}_F}{\hat{\beta}_R - \hat{\beta}_C},
$$
The statistic $\delta$ is interpreted as the magnitude of covariance between the unobserved part of the controls and the treatment variable necessary to explain away the entire treatment effect of $X$ on $Y$. A larger ratio suggests it is implausible that omitted variable bias could explain away the entire observed effect. See @BellowsMiguel2009a [Appendix A] for the derivation.
@NunnWantchekon2011a provides a clear explanation and application  of the  statistic. 

Often you will see works that add regressors sequentially and make some sort of implicit coefficient stability argument. That is not useful. The important comparison is between the coefficient when nothing (or only a small subset of covariates) is controlled for, and the full set of controls.

@BellowsMiguel2009a themselves generalize @AltonjiElderTaber2005a from binary to continuous treatment variables. @Oster2016a further generalizes the estimator. @PeiPischkeSchwandt2017a show that if the covariates are measured with error, a "balancing test" (regressing the confounder on the treatment) is
more powerful.  

Methods such as *matching*, *propensity scores*, or *inverse weighting* still depend on assumptions about selection on observables. 
They may be less sensitive to "omitted variable bias" due to 
The differ from regression in the estimand or their sensitivity to model misspecification.

The preference for "design based" inference is mostly driven by a desire to find situations (designs) where other assumptions can substitute for the nigh impossible to test "selection on observables" assumption. Apart from experiments, these include instrumental variables, regression discontinuity, and difference-in-differences.


## Measurement Error

### What's the problem?

It biases coefficients:

1. Variable with measurement error: biases $\beta$ towards zero (**attenuation bias**)
2. Other variables: Biases $\beta$ similarly to omitted variable bias. In other words, when a variable has measurement error it is an imperfect control. You can think of omitted variables as the limit of the effect of measurement error as it increases.


### What to do about it?

There's no easy fix within the OLS framework.

1. If the measurement error is in the variable of interest, then the variable will be biased towards zero, and your estimate is too large.
2. Find better measures with lower measurement errors. If the variable is the variable of interest, then perhaps combine multiple variables into a single index. If the measurement error is in the control variables, then include several measures. That these measure correlate closely increases their standard errors, but the control variables are not the object of the inferential analysis.
3. More complicated methods: errors in variable models, structural equation models, instrumental variable (IV) models, and Bayesian methods.

@BlackwellHonakerKing2015a note that the easiest way to handle measurement error in the predictors is to treat them as missing data where you have extra information about their range.  Suppose a covariate is observed as $x \sim(x^*, \delta^2)$, where $x^*$ is the true value, and $\delta$ is the scale of the measurement error. Then missing values are the case when $\delta \to \infty$. 
So missing values are special, extreme, case of measurement error. 
This means that we can use multiple imputation methods for dealing with [missing values](missing_values.html) where we add additional information to restrict the plausible range of imputated values. The `r rpkg("Amelia")` has built-in support for this, but the general idea could be adapted to other multiple imputation methods.



## More Information

### Simpson's Paradox

- See @Samuels1993a for more discussion of Simpson's Paradox
- @Moore2005a collects and succinctly describes several examples of Simpson's Paradox
- An interactive visualization of the [Simpson's Paradox](http://vudlab.com/simpsons/)
- Horton. 2015. [Fun with Simpson's Paradox: Simulating Confounders](http://blog.revolutionanalytics.com/2015/11/fun-with-simpsons-paradox-simulating-confounders.html)
- Horton. 2012. [Example 9.20: visualizing Simpson’s paradox](https://www.r-bloggers.com/example-9-20-visualizing-simpsons-paradox/)
- See the [Wikipedia Page](https://en.wikipedia.org/wiki/Simpson%27s_paradox)
- [US Median Wage by Education Level](https://economix.blogs.nytimes.com/2013/05/01/can-every-group-be-worse-than-average-yes/). Overall wages have risen, but within every group, the wage has fallen.
- Nielsen. [Reinventing Explanation](http://michaelnielsen.org/reinventing_explanation/index.html) has a visual explanation of the Simpson's paradox
- Gelman. [Understanding Simpson's Praradox Using a Graph](http://andrewgelman.com/2014/04/08/understanding-simpsons-paradox-using-graph/). April 8, 2014. Discusses the Nielsen post, provides other visualizations, and notes how aggregation problems arise even in non-causal cases.
- @ArmstrongWattenberg2014a introduce the Comet Chart for visualizing Simpson's Paradoxes.
  See this [page](https://www.zanarmstrong.com/#/research-1/) for code and examples, including an [R implementation](https://gist.github.com/zanarmstrong/6c2855a34f504029847485c690692e75).
