---
title: "Regression.Rmd"
author: "Jeffrey Arnold"
date: "3/31/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}
library("datums")
library("tidyverse")
filter <- dplyr::filter
```

# Pauperism Data

```{r}
pauperism_year %>%
  filter(year == 1871) %>%  
  select(outratio, pauper2) %>%
  drop_na() %>%
  ggplot(aes(x = outratio, pauper2)) +
  geom_point() + 
  geom_smooth(method = "lm")
```

```{r}
pauperism_year %>%
  filter(year == 1871) %>%  
  select(outratio, pauper2) %>%
  drop_na() %>%
  cor()
```

Regressions for each year
```{r}
pauperism_year %>%
  group_by(year) %>%
  summarise(mod = list(lm(outratio ~ pauper2, data = .)))
```


# Anscombe quartet

```{r}
anscombe_tidy <-
  anscombe %>%
  mutate(obs = row_number()) %>%
  gather(variable_dataset, value, -obs) %>%
  separate(variable_dataset, c("variable", "dataset"), sep = c(1)) %>%
  spread(variable, value) %>%
  arrange(dataset, obs)

```

What are summary statistics of the four anscombe datasets?
```{r}
ggplot(anscombe_tidy, aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ dataset, ncol = 2)
```

What are the mean, standard deviation, correlation coefficient, and regression
coefficients of each line? 
```{r}
anscombe_summ <-
  anscombe_tidy %>%
  group_by(dataset) %>%
  summarise(
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x = sd(x),
    sd_y = sd(y),
    cov = cov(x, y),
    cor = cor(x, y),
    coefs = list(coef(lm(y ~ x, data = .)))
  ) %>%
  mutate(
    intercept = map_dbl(coefs, "(Intercept)"),
    slope = map_dbl(coefs, "x")
  ) %>%
  select(-coefs)

```

WUT? They are the same? But they look so different. Of course that was the point ...

Since this all revolves around covariance, lets calculate the values of $x_i - \bar{x}$,
$y_i - \bar{y}$, and $(x_i - \bar{x}) (y_i - \bar{y})$ for each obs for each variable.
```{r}
anscombe_tidy <-
  anscombe_tidy %>%
  group_by(dataset) %>%
  mutate(mean_x = mean(x),
         diff_mean_x = x - mean_x,
         mean_y = mean(y),
         diff_mean_y = y - mean_y,
         diff_mean_xy = diff_mean_x * diff_mean_y,
         quadrant = 
           if_else(
             diff_mean_x > 0, 
             if_else(diff_mean_y > 0, 1, 2),
             if_else(diff_mean_y > 0, 4, 3),
           ))
```

```{r}
ggplot(anscombe_tidy, aes(x = x, y = y,
                          size = abs(diff_mean_xy),
                          colour = factor(sign(diff_mean_xy)))) +
  geom_point() +
  geom_hline(data = anscombe_summ,
             aes(yintercept = mean_y)) +
  geom_vline(data = anscombe_summ,
             aes(xintercept = mean_x)) +  
  facet_wrap(~ dataset, ncol = 2)
```

```{r}
ggplot(anscombe_tidy, aes(x = x, y = y, colour = factor(sign(diff_mean_xy)))) +
  geom_point() +
  geom_segment(mapping = aes(xend = mean_x, yend = mean_y)) +
  geom_hline(data = anscombe_summ,
             aes(yintercept = mean_y)) +
  geom_vline(data = anscombe_summ,
             aes(xintercept = mean_x)) + 
  facet_wrap(~ dataset, ncol = 2)
```

```{r}
ggplot(anscombe_tidy, aes(x = x, y = y, colour = factor(sign(diff_mean_xy)))) +
  geom_point() +
  geom_segment(mapping = aes(xend = x, yend = mean_y)) +
  geom_segment(mapping = aes(xend = mean_x, yend = y)) +  
  geom_hline(data = anscombe_summ,
             aes(yintercept = mean_y)) +
  geom_vline(data = anscombe_summ,
             aes(xintercept = mean_x)) +  
  facet_wrap(~ dataset, ncol = 2)

```

# What is regression? 

Different conditional means ... 

Use pauperism data


# Linear Regression 

Objective function and weights


```{r}
pauperism_year %>%
  filter(year == 1871) %>%
  select(outratio, pauper2) %>%
  drop_na() %>%
  lm(pauper2 ~ outratio, data = .)
```

```{r}

```

