# Simpson's Paradox

[This](http://vudlab.com/simpsons/) is good visualization of Simpson's Paradox (except the subtitle).


E.g. https://www.youtube.com/watch?v=ebEkn-BiW5k

## Example


### Batting Averages

>  Ken Ross. "A Mathematician at the Ballpark: Odds and Probabilities for Baseball Fans (Paperback)" Pi Press, 2004. ISBN 0-13-147990-3. 12–13

```{r}
batting_yearly <- tribble(
  ~ player, ~ year, ~ hits, ~ ab,
  "Derek Jeter", 1995, 12, 48,
  "Derek Jeter", 1996, 183, 582,  
  "David Justice", 1995, 104, 411,
  "David Justice", 1996, 45, 140
) %>%
  mutate(avg = hits / ab)
```

```{r}
batting_total <- batting_yearly %>%
  group_by(player) %>%
  summarise(ab = sum(ab), hits = sum(hits)) %>%
  mutate(avg = hits / ab)
```

```{r}
ggplot() +
  geom_point(data = batting_yearly, 
             mapping = aes(x = as.integer(as.factor(player)),
                           y = avg, 
                           color = as.factor(year), 
                           size = ab)) +     
  geom_line(data = batting_yearly,
             mapping = aes(x = as.integer(as.factor(player)),
                           y = avg, color = as.factor(year))) +
  geom_point(data = batting_total, 
             mapping = aes(x = as.integer(as.factor(player)),
                           y = avg, 
                           size = ab)) +
  geom_line(data = batting_total,
             mapping = aes(x = as.integer(as.factor(player)),
                           y = avg)) +
  scale_x_continuous("Player", breaks = 1:2, 
                     labels = levels(as.factor(batting_yearly$player))) +
  scale_y_continuous("Batting Avg.") +
  scale_color_discrete("Year") +
  scale_size_continuous("At Bats")

```

### Kidney Stones

See https://doi.org/10.1136/bmj.309.6967.1480

```{r}
kidney_stones <- tribble(
  ~ treatment, ~ size, ~ success, ~ n,
  0, "Small", 81, 87,
  0, "Large", 192, 263,
  1, "Small", 234, 270,
  1, "Large", 55, 80
) %>%
  mutate(p = success / n)
```

```{r}
kidney_stones_total <- kidney_stones %>%
  group_by(treatment) %>%
  summarise(success = sum(success), n = sum(n)) %>%
  mutate(p = success / n)
```

```{r}
ggplot() +
  geom_point(data = kidney_stones, 
             mapping = aes(x = treatment, y = p, 
                           color = size, size = n)) +     
  geom_line(data = kidney_stones, 
             mapping = aes(x = treatment, y = p, color = size)) +
  geom_point(data = kidney_stones_total, 
              mapping = aes(x = treatment, y = p, size = n)) +
  geom_line(data = kidney_stones_total,
            mapping = aes(x = treatment, y = p)) +
  scale_x_continuous("Treatment", 
                     breaks = c(0, 1), labels = c("A", "B")) +
  scale_y_continuous("% Success") +
  scale_color_discrete("Kidney Stone Size") +
  scale_size_continuous("Number in Group")

```

### Blood Pressure Drug

Example from Pearl et al. (p. 4) originall from Simpson (1951).

Consider data on 700 sick patients given the opportunity to try a new drug, of which 350 *chose* to take the new drug.
The number of patients in each gender ("Male", "Female") and the number recovered were recorded.

```{r}
tribble(
  ~ gender, ~ drug, ~ recovered, ~ n,
  "Men", 1, 81, 87,
  "Men", 0, 234, 270,
  "Women", 1, 192, 263,
  "Women", 0, 289, 350
) 
```

1. What is the overall recovery rate ? 
2. What is the recovery rate within each gender?


### Berkeley Admissions

Data in 

See visualizations in https://www.ft.com/content/94e3acec-a767-11e7-ab55-27219df83c97
and 


### Teacher Salary and Test Scores

> Deborah Lynn Guber, "Getting what you pay for: the debate over equity in public school expenditures" (1999), *Journal of Statistics Education*

```{r}
data("SAT", package = "mosaicData")
```

```{r}
ggplot(SAT, aes(label = state, x = salary, y = sat)) +
  geom_text() +
  geom_smooth(method = "lm", se = FALSE)
```


```{r}
mutate(SAT,
       frac_cat = cut_number(frac, 3)) %>%
  ggplot(aes(label = state, x = salary, y = sat, color = frac_cat)) +
  geom_text() +
  geom_smooth(method = "lm", se = FALSE)
```


### Other Examples

- U.S. Education and Income. Norris, Floyd. "[Can Every Group Be Worse Than Average? Yes.](https://economix.blogs.nytimes.com/2013/05/01/can-every-group-be-worse-than-average-yes/)". *New York*. May 1, 2013. 
- Armstrong, Zan, and Martin Mattenberg. 2014. "[Visualizing Statistical Mix Effects and Simpson’s Paradox](https://static.googleusercontent.com/media/research.google.com/en//pubs/archive/42901.pdf)"
- Horton, Bob. [Fun with Simpson's Paradox: Simulating Confounders](http://blog.revolutionanalytics.com/2015/11/fun-with-simpsons-paradox-simulating-confounders.html) November 17, 2015.
- Burn-Murdoch, John. "[Germany’s election and the trouble with correlation](https://www.ft.com/content/94e3acec-a767-11e7-ab55-27219df83c97)" *Financial Times*. October 2, 2017.
- Income and Party Affiliation in the United States. DOI 10.1561/100.00006026
