# Bivariate OLS

## Regression Line

The regression line is computed from five summary statistics

- mean of $x$
- mean of $y$
- standard deviation of $x$
- standard deviation of $y$
- correlation of $x$ and $y$

The regression line is flatter for the standard deviation line

```{r}
filter <- dplyr::filter
library("HistData")
heights <- dplyr::filter(PearsonLee, par == "Father", chl == "Son")
heights %>%
  ggplot(aes(x = parent, y = child)) + 
  geom_jitter()
       
```

```{r}
height_summary <-
  heights %>%
  group_by(parent) %>%
  summarise(child = mean(child))

ggplot() +
  geom_jitter(data = heights, aes(x = parent, y = child), alpha = 0.3) +
  geom_smooth(data = heights, aes(x = parent, y = child), se = FALSE, method = "lm", colour = "red") +
  geom_point(data = height_summary, aes(x = parent, y = child), colour = "black", size = 2)

```

"Corelation measures the extent to which a scatter diagram is packed around a line" - Freedman p. 21

$$
MSE  = (1 - r^2) \Var(y)
$$

Among all lines, the regression line has the smallest RMSE

- estimates aren't parameters, and residuals aren't random errors
- in a regression model, the data are observed values of random variables.
- observed values are called "realizations"

```{r}
heights %>%
  add_predictions(lm(parent ~ child, data = .), var = "pred_parent") %>%
  add_predictions(lm(child ~ parent, data = .), var = "pred_child") %>%
  mutate(sd_y = sd(child),
         sd_x = sd(parent),
         mean_y = mean(child),
         mean_x = mean(parent),
         cor_xy = cor(parent, child),
         cor_line = mean_y + sign(cor_xy) * (sd_y / sd_x) * (parent - mean_x)) %>%
  ggplot() +
  geom_jitter(mapping = aes(x = parent, y = child), alpha = 0.3) +
  geom_line(mapping = aes(x = parent, y = pred_child), colour = "blue") +
  geom_line(mapping = aes(x = pred_parent, y = child), colour = "red") +
  geom_line(mapping = aes(x = parent, y = cor_line))


```

The correlation line is
$$
y = \bar{y} + \text{sign}(r_{xy}) \left( \frac{s_y}{s_x} \right) (x - \bar{x})
$$



## OLS as the Weighted Sum of Outcomes


```{r}

```



