# Simulation examples

```{r}
## ----sdcor2cov-----------------------------------------------------------
sdcor2cov <- function(s, r = diag(length(s))) {
  s <- diag(s, nrow = length(s), ncol = length(s))
  s %*% r %*% s
}

## ----sim_lin_norm--------------------------------------------------------
sim_lin_norm <- function(iter, n, mu_X, s_X, R_X, beta, sigma) {
  # Error checking so that bugs are caught quicker :-)
  assert_that(length(s_X) == length(mu_X),
              ncol(R_X) == nrow(R_X),
              ncol(R_X) == length(mu_X),
              length(beta) == (length(mu_X) + 1))
  # Generate an X
  X <- MASS::mvrnorm(n, mu = mu_X, Sigma = sdcor2cov(s_X, R_X),
                     empirical = TRUE)
  # Create a list to stor the results
  iterations <- list()
  # Create a progress bar because we're impatient
  p <- progress_estimated(iter, min_time = 2)
  # Loop over the simulation runs
  for (j in 1:iter) {
    # Draw y
    mu <- cbind(1, X) %*% beta
    epsilon <- rnorm(n, mean = 0, sd = sigma)
    y <- mu + epsilon
    # Run a regression
    mod <- lm(y ~ X)
    # Save the coefficients in a data frame
    mod_df <- tidy(mod) %>%
      # Add a column indicating the simulation number
      mutate(.iter = j)
    # Add hetroskedasticity consistent se to the data
    mod_df[["std.error.robust"]] <- sqrt(diag(car::hccm(mod)))
    # Save these results as the next element in the storage list
    iterations[[j]] <- mod_df
    # Update the progress bar
    p$tick()$print()
  }
  # Combine the list of data frames into a single data frame
  bind_rows(iterations)
}


## ----summarize_sim-------------------------------------------------------
summarize_sim <- function(.data, beta) {
  ret <- .data %>%
    group_by(term) %>%
    summarize(estimate_mean = mean(estimate),
              estimate_sd = sd(estimate),
              se_mean = sqrt(mean(std.error) ^ 2),
              se_robust_mean = sqrt(mean(std.error.robust) ^ 2),
              iter = length(estimate))
  ret[["beta_true"]] <- beta
  ret
}

## ----sim_lin_norm_omitted------------------------------------------------
sim_lin_norm_omitted <- function(iter, n, mu_X, s_X, R_X, beta, sigma,
                                 omit = integer(0)) {
  assert_that(length(s_X) == length(mu_X),
              ncol(R_X) == nrow(R_X),
              ncol(R_X) == length(mu_X),
              length(beta) == (length(mu_X) + 1))
  # Generate an X
  k <- length(mu_X)
  X <- MASS::mvrnorm(n, mu_X, sdcor2cov(s_X, R_X), empirical = TRUE)
  # ------
  # NEW: ensure colnames of X are consistent despite omitting some in lm
  colnames(X) <- paste("X", 1:k, sep = "")
  # ------
  iterations <- list()
  p <- progress_estimated(iter, min_time = 2)
  for (j in 1:iter) {
    mu <- cbind(1, X) %*% beta
    epsilon <- rnorm(n, mean = 0, sd = sigma)
    y <- mu + epsilon
    # ---------
    # NEW: omit columns of X
    # Look up paste and setdiff function to see what they does
    Xomit <- as.data.frame(X)[ , setdiff(1:k, omit)]
    # ~ . means use all variables from `data` on the RHS of the formula
    mod <- lm(y ~ . , data = Xomit)
    # ---------
    mod_df <- tidy(mod) %>%
      mutate(.iter = j)
    mod_df[["std.error.robust"]] <- sqrt(diag(car::hccm(mod)))
    iterations[[j]] <- mod_df
    p$tick()$print()
  }
  # Combine the list of data frames into a single data frame
  bind_rows(iterations)
}

## ----sim_lin_norm_heterosked-----------
sim_lin_norm_heterosked <- function(iter, n, mu_X, s_X, R_X, beta, gamma) {
  assert_that(length(s_X) == length(mu_X),
              ncol(R_X) == nrow(R_X),
              ncol(R_X) == length(mu_X),
              length(beta) == (length(mu_X) + 1),
              length(gamma) == (length(mu_X) + 1))
  X <- MASS::mvrnorm(n, mu_X, sdcor2cov(s_X, R_X), empirical = TRUE)
  iterations <- list()
  p <- progress_estimated(iter, min_time = 2)
  for (j in 1:iter) {
    mu <- cbind(1, X) %*% beta
    # ------------
    # NEW: variance varies by each observation
    sigma <- sqrt(exp(cbind(1, X) %*% gamma))
    # ------------
    epsilon <- rnorm(n, mean = 0, sd = sigma)
    y <- mu + epsilon
    # Run a regression
    mod <- lm(y ~ X)
    # Save the coefficients in a data frame
    # and Add a column indicating the simulation number
    mod_df <- tidy(mod) %>%
      mutate(.iter = j)
    mod_df[["std.error.robust"]] <- sqrt(diag(car::hccm(mod)))
    iterations[[j]] <- mod_df
    p$tick()$print()
  }
  bind_rows(iterations)
}

## ----sample_lin_norm_heterosked-----------
sample_lin_norm_heterosked <- function(iter, n, mu_X, s_X, R_X, beta, gamma) {
  assert_that(length(s_X) == length(mu_X),
              ncol(R_X) == nrow(R_X),
              ncol(R_X) == length(mu_X),
              length(beta) == (length(mu_X) + 1),
              length(gamma) == (length(mu_X) + 1))
  X <- MASS::mvrnorm(n, mu_X, sdcor2cov(s_X, R_X), empirical = TRUE)
  iterations <- list()
  p <- progress_estimated(iter, min_time = 2)
  for (j in 1:iter) {
    mu <- cbind(1, X) %*% beta
    # ------------
    # NEW: variance varies by each observation
    sigma <- sqrt(exp(cbind(1, X) %*% gamma))
    # ------------
    epsilon <- rnorm(n, mean = 0, sd = sigma)
    y <- as.numeric(mu + epsilon)
    iterations[[j]] <-
      data_frame(y = y) %>%
      bind_cols(as.data.frame(X)) %>%
      mutate(.iter = j)
    p$tick()$print()
  }
  bind_rows(iterations)
}

## ----sim_lin_norm_truncated----------------------------------------------
sim_lin_norm_truncated <- function(iter, n, mu_X, s_X, R_X, beta, sigma,
                                   truncation = 0.5) {
  X <- MASS::mvrnorm(n, mu_X, sdcor2cov(s_X, R_X), empirical = TRUE)
  iterations <- list()
  p <- progress_estimated(iter, min_time = 2)
  for (j in 1:iter) {
    mu <- cbind(1, X) %*% beta
    epsilon <- rnorm(n, mean = 0, sd = sigma)
    y <- mu + epsilon
    # -------
    # NEW: drop cases in which y > mean(y)
    is_obs <- (y > quantile(y, prob = truncation))
    yobs <- y[is_obs, ]
    Xobs <- X[is_obs, ]
    # give consistent colnames
    colnames(Xobs) <- paste("X", 1:k, sep = "")
    colnames(yobs) <- "y"
    # ------
    # -------
    # Run a regression
    mod <- lm(yobs ~ Xobs)
    # Save the coefficients in a data frame
    # and Add a column indicating the simulation number
    mod_df <- tidy(mod) %>%
      mutate(.iter = j)
    mod_df[["std.error.robust"]] <- sqrt(diag(car::hccm(mod)))
    iterations[[j]] <- mod_df
    p$tick()$print()
  }
  bind_rows(iterations)
}
```



---
title: "POLS 503: Problem Set 3"
author: "Jeffrey B. Arnold, Christopher Adolph"
date: "April 30, 2015"
---

*Revised:* May 8, 2015

$$
\DeclareMathOperator{\cor}{cor}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\sd}{sd}
\DeclareMathOperator{\se}{se}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\quantile}{quantile}
$$


```{r knitr_opts, echo = FALSE, message = FALSE, results = 'hide'}
library("knitr")
set.seed(135125)
opts_chunk$set(cache = TRUE, autodep = TRUE, cache.extra = rand_seed)
```
```{r load,message = FALSE, echo = FALSE}
library("MASS")
library("ggplot2")
library("dplyr")
library("broom")
library("assertthat")
library("tidyr")
```
```{r external,cache = FALSE, echo=FALSE}
read_chunk("hw3-functions.R")
```
```{r echo=FALSE}
select <- dplyr::select
```

# Instructions and Introduction

The purpose of this homework is to provide a guided, hands-on tour through the properties of the least squares estimator, especially under common violations of the Gauss Markov assumptions.
We will work through a series of programs which use simulated data --- i.e., data created with known properties --- to investigate how these violations affect the accuracy and precision of least squares estimates of slope parameters.
Using repeated study of simulated datasets to explore the properties of statistical models is called Monte Carlo experimentation.
Although you will not have to write much R code, you will need to read through
the provided programs carefully to understand what is happening.

Monte Carlo experiments always produce the same results as analytic proofs for the specific case considered.
Each method has advantages and disadvantages: proofs are more general and elegant, but are not always possible.
Monte Carlo experiments are much easier to construct and can always be carried out, but findings from these experiments only apply to the specific scenario under study.
Where proofs are available, they are generally preferable to Monte Carlo experiments, but proofs of the properties of more complicated models are sometimes impossible or impractically difficult.
This is almost always the case for the properties of models applied to small samples of data.
Here, we use Monte Carlo not out of necessity but for pedagogical purposes, as a tool to gain a more intuitive and hands-on understanding of least squares and its properties.

1. Create a new R project for this homework named `hw3` and load that project.
2. Do your analyses in an R markdown document based off of [hw3-template.Rmd](hw3-template.Rmd)
2. Download [hw3-functions.R](hw3-functions.R) and save in to your project directory.
3. Submit a zipped file of the directory with your R project through Canvas. This should contain all the materials for another person to run your R Markdown file  This should contain:

    - The R project (`.Rproj`) file
	  - The R Markdown document (`.Rmd`) of your analyses
	  - An HTML document (`.html`) compiled from your R Markdown document.
	  - Any data or other files needed to run the analyses in your R Markdown document.

4. Your R Markdown file should follow the guidance from [here](http://pols503.github.io/pols_503_sp15/r_markdown_assignments.html) and your R code should follow the guidelines [here](http://pols503.github.io/pols_503_sp15/r_best_practices.html).
5. Turn in a paper copy of the document compiled from the analyses at lab.
6. You can work together on this but you should each turn in your own assignments and write up your work separately.
   Include the names of your collaborators on your assignment.

# Setup

This will use the standard [Hadleyverse](http://blog.revolutionanalytics.com/2015/03/hadleyverse.html) packages that we've been using in this course (**ggplot2**, **dplyr**, **tidyr**, **broom**).
A few of the functions will use **assertthat**, which contains functions to test for errors in functions.
Additionally we will use the `hccm` function from **car**, but we will not load **car** since it contains some function names that conflict with those in packages that we are using.
```{r eval = FALSE, echo = TRUE}
<<load>>
```

# Simulation Example

All of the simulations in this assignment will follow the same structure:

1. Define a population
2. Repeat $m$ times:

    1. Draw a sample from the population
    2. Run OLS on that sample
    3. Save statistics, e.g. coefficients, standard errors, $p$-values, from the sample regression.

3. Evaluate the distributions of the sample statistics, or summaries thereof, to determine how well OLS recovers the parameters of the population.

In this section, we will work through the code necessary to run a simulation.
However, in the problems, functions written for this problem set will do most of the simulation computation. 
This section is to help you to understand what those functions are doing, and to provide a mapping from the math to the code.

We will use an example in which the population satisfies all the Gauss-Markov assumptions and we run a correctly specified regression on the samples drawn from that population.

In this example, the population model is
$$
\begin{aligned}[t]
Y_i &= \beta_0 + \sum_{j = 1}^k \beta_j x_{i,j} + \epsilon_i \\
\epsilon_i & \sim N(0, \sigma^2)
\end{aligned}
$$
For a sample $y$ from that population, the OLS regression which will be run is
$$
\begin{aligned}[t]
y_i &= \hat\beta_0 + \sum_{j = 1}^k \hat\beta_j x_{i,j} + \hat\epsilon_i \\
\hat\sigma^2 &= \frac{\sum \hat\epsilon_i }{n - k - 1}
\end{aligned}
$$
In this case, the regression run on the samples has the correct specification, but that will not necessarily be true for other examples.

In this section, we will proceed in two steps.

1. Write code to generate a single sample and run OLS on it.
2. Generalize that code by

    a. Putting it in a loop to be able to draw many samples
    b. Putting it in a function to make it easy to change parameters of the simulation.

## Single Iteration

### Drawing $X$

First, we need to generate some values of $X$ that we will use in the samples.
Recall that the sampling distributions of OLS coefficients and the Gauss-Markov theorem are defined for a fixed $X$.[^randomx]
So, we will randomly generate data for the covariates, but use the same values of the covariates for all samples from $Y$.
Although linear regression does not require covariates to be distributed multivariate normal, we will generate $X$ by drawing a sample of size $n$ from a multivariate normal distribution with mean $\mu_X$ and covariance matrix $\Sigma_X$.
$$
X_i \sim N(\mu_X, \Sigma_X) \text{ for $i = 1, \dots, n$.}
$$
Since covariance matrices are not particularly intuitive, so it may be easier to decompose the covariance into a correlation matrix and the standard deviations of the variables.
A $k \times k$ covariance matrix, $\Sigma$ can be decomposed into a vector of $k$ standard deviations, $s$, and a $k \times k$ correlation matrix, $R$:
$$
\begin{aligned}[t]
\Sigma &= S R S &
&& S
& = \diag(s)
= 
\begin{bmatrix}
s_1 & 0 & \dots & 0 \\
0 & s_2 & \dots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \dots & s_k
\end{bmatrix}
\end{aligned}
$$
The function `sdcor2cov` in `hw3-functions.R` calculates the covariance matrix from a standard deviation
and a correlation matrix.
```{r sdcor2cov, echo=FALSE}
```

In this example, we will use a sample of size $n = 1024$, with $k = 3$ independent variables drawn from a multivariate normal distribution with mean $\mu_X = (0, 0, 0)$, standard deviations of $s_X = (1, 1, 1)$.
If variables are independent, then their correlation matrix is the identity matrix (a diagonal matrix with all 1's on the diagonal):
$$
R_X = \begin{bmatrix} 1 & 0 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}
$$
Since the variables are independent, this is equivalent to separately sampling each variable from a standard normal distribution.
```{r}
n <- 1024
mu_X <- rep(0, 3)
s_X <- rep(1, 3)
R_X <- diag(3)
Sigma_X <- sdcor2cov(s_X, R_X)
```
We draw the sample using the **MASS** function `mvrnorm`,
```{r}
X <- mvrnorm(n, mu_X, Sigma_X, empirical = TRUE)
```
The option `empirical = TRUE` is used to make sure that although $X$ is randomly sampled,
it is adjusted to so that the sample mean and covariance are equal to $\mu_X$ and $\Sigma_X$,
```{r}
round(cor(X), 1)
round(apply(X, 2, mean), 1)
```

### Drawing Y

After defining $X$, we need values of $\beta$ and $\sigma$ to draw samples from $Y$.

For this example, set the true parameters of the model so that the intercept is 0, and the slope coefficients are all 1.
```{r}
beta <- c(0, 1, 1, 1)
```
and set the standard deviation of the regression errors such that that the $R^2$ of the regression is approximately 0.5
```{r}
sigma <- 1.7
```
Calculate the expected value of the outcome conditional on the covariates, $E(Y | X)$,
```{r}
mu_y <- cbind(1, X) %*% beta
```
The expression `cbind(1, X)` adds a column of 1s as an intercept in the regression to the covariates in $X$.
Then sample the errors, $\epsilon \sim N(0, \sigma^2)$,
```{r}
epsilon_y <- rnorm(n, mean = 0, sd = sigma)
```
Now combine the systematic component, $E(Y | X)$, and stochastic component, $\epsilon$, to generate the values of $y$ in the sample,
```{r}
y <- mu_y + epsilon_y
```

### Sample Regression

Now that we have a sample, we will run an OLS regression on it in order to estimate the parameters of the population,
```{r}
mod <- lm(y ~ X)
```
We will use the `tidy` function from the **broom** package convert the coefficients, standard errors, p-values, and t-values into a data frame.
This will be especially useful when storing the results from many simulations.
```{r}
mod_df <- tidy(mod)
```
The function `tidy` does not include a heteroskedasticity consistent standard error, and we would like to compare that to the classical standard error in some simulations.
We will use the function `hccm` from the **car** package to calculate the heteroskedasticity robust standard error.
```{r}
mod_df[["std.error.robust"]] <- sqrt(diag(car::hccm(mod)))
```

The coefficients of the OLS regression on the parameter should be similar to, but not exactly, those of the population from which it was drawn.
```{r}
mod_df
```

## Multiple iterations

We want to repeat this many times in order to generate a sampling distribution of statistics of interest in order to evaluate how well they work as estimators.
We will do this by (1) wrapping the code from the previous section in a for loop, so we can repeat the simulations many times, and (2) put it all in a function, so that we can easily change the inputs.

Thus, we will define a function, named `sim_lin_norm`.
The code for this function is in the chunk below.
The function itself takes several arguments

`iter`

:    The number of iterations to run / samples to draw.

`n`

:    The number of observations

`mu_X`

:    The means of the covariates in $X$. The number of covariates do not need to
     be specified as it is inferred from the length of `mu_X`.

`s_X`

:    The standard deviations of the covariates in $X$.

`R_X`

:    The correlation matrix for the covariates in $X$.

`beta`

:    The coefficient vector, $\beta$.

`sigma`

:    The standard deviations of the regression errors, $\sigma$.


This function returns a data frame with the results of all the simulations.
Each row is a coefficient (column `term`) for a simulation (column `.iter`)

```{r sim_lin_norm}
```
Now we will draw samples using the values that we used before and look at the results.
```{r sim0}
iter <- 1024
sim0 <- sim_lin_norm(iter, n, mu_X, s_X, R_X, beta, sigma)
head(sim0)
```
With all the samples, we can summarize the results and compare them to the population parameters to evaluate how well OLS works.
The following function, `summarize_sim` takes a data frame generated by the simulation function and a vector of the original parameters, and generates a summary data frame with one row per coefficient.
Its arguments are

`.data`

:    A data frame of a simulation produced by one of the functions in this problem set.

`beta`

:    The true values of the population parameters $\beta$.

```{r summarize_sim}
```
Using the previous results, the summary is
```{r sim0_summary}
sim0_summary <- summarize_sim(sim0, beta)
sim0_summary
```

In these simulations we will generally, but not exclusively, be concerned with the bias and efficiency of the estimates.

First, compare the average value of $\hat\beta$ to the population parameters $\beta$.
This is what is nice about simulations, we know what the correct answer should be!
```{r}
select(sim0_summary, estimate_mean, beta_true) %>%
  mutate(bias = estimate_mean - beta_true)
```
If an estimator is unbiased, then the mean of its sampling distribution should be equal to the true parameter value.
Note that these may not be exactly the same due to the randomness from taking a sample; this is called Monte Carlo error.

Second, we are interested in the standard deviation of the $\hat\beta$ estimates,
We will often be interested in how the standard deviation of the estimator changes with inputs to the simulation.
```{r}
select(sim0_summary, estimate_sd)
```

Third, we will be interested in whether the standard error, which is sample estimate of the standard deviation of the sampling distribution, is a good estimate of the actual sampling distribution of $\hat\beta$.
The mean of the classical standard errors of the simulations are in column `se_mean`.
The column `se_robust_mean` contains the means of the heteroskedasticity robust standard errors.
```{r}
select(sim0_summary, estimate_sd, se_mean) %>%
  mutate(bias = se_mean - estimate_sd)
select(sim0_summary, estimate_sd, se_robust_mean) %>%
  mutate(bias = se_robust_mean - estimate_sd)
```


If `se_mean` (`se_robust_mean`) is not equal (or very close) to the `estimate_sd`, then the standard errors (robust standard errors) calculated from samples are biased estimates of the standard deviation of the sampling distribution of $\hat{\beta}$.

Those are the most common questions we will consider in the simulations,
but some simulations may consider other questions.

See the first problem for an example of how to put this all together.

# Problems {#problems}

The problems of this assignment will use various simulations similar to the one in the previous section.
For all simulations use `iter = 1024`.
If this is taking too long you can reduce the number of iterations.

```{r}
iter <- 1024
```


## Linear Normal Model with Homoskedasticity

In this problem, we will use `sim_lin_norm` to understand how OLS recovers parameters when the population meets the Gauss-Markov assumptions.
$$
\begin{aligned}[t]
Y_i &= 0 + 1 \cdot x_{1,i} + 1 \cdot x_{2,i} + 1 \cdot x_{3,i} + \epsilon_i \\
\epsilon_i &\sim N(0, \sigma^2) \\
\sigma &= 1.7
\end{aligned}
$$
The covariates, $X$, are independent, with means, $\mu_{X,j} = 0$, and standard deviations, $s_{X,j} = 1$ for all variables.
The sample size $n$ will vary with simulation.

You would want to set these variables to use in a simulation
```{r}
mu_X <- c(0, 0, 0)
s_X <- c(1, 1, 1)
R_X <- diag(3)
beta <- c(0, 1, 1, 1)
sigma <- 1.7
```

Run several simulations with `sim_lin_norm` varying the sample size to determine how sample size affects the following: $n = \{32, 64, 512, 1024 \}$.
We will store these in a vector to use later:
```{r}
n_samples <- c(32, 64, 512, 1024)
```

As shown in the introduction, you can use `sim_lin_norm` to simulate for a single sample size as follows.
E.g. for sample size 32, we could run
```{r}
n <- 32
summarize_sim(sim_lin_norm(iter = iter, n = n,
                           mu_X = mu_X, s_X = s_X, R_X = R_X,
                           beta = beta, sigma = sigma),
              beta = beta)
```
In order to run this for *all* the different sample sizes, we will put this in a list
```{r tidy = FALSE}
sims1 <- list()
# Loop over the integers 1 to the number of different sample sizes.
for (i in 1:length(n_samples)) {
  # Run simulation
  sim_results <- sim_lin_norm(iter, n_samples[i], mu_X, s_X, R_X, beta, sigma)
  # Summarize simmulations
  sim_summ <- summarize_sim(sim_results, beta)
  # Add a new column to the data to indicate which sample size this simulation used
  sim_summ[["n"]] <- n_samples[i]
  # Save the data to the list in location i
  sims1[[i]] <- sim_summ
}
sims1 <- bind_rows(sims1)
```

Then use these results to analyze the following:

- The bias of each $\hat{\beta}_j$. How different is `estimate_mean` ($\hat\beta$) from `beta_true` ($\beta$)?
- The standard deviation of $\hat{\beta}_j$?. How does `estimate_sd` change as the sample size increases?
- The bias of the standard error of each $\hat{\beta}_j$. Compare the actual population standard deviations, `estimate_sd`, with the means of the standard errors, `se_mean`.
- The bias of the robust standard error of each $\hat{\beta}_j$? Compare the actual population standard deviations of each parameter, `estimate_sd`, with the means of the robust standard errors, `se_robust_mean`.

As an example, to find the bias of $\hat\beta$:
```{r}
sims1 %>%
  mutate(bias = estimate_mean - beta_true) %>%
  select(term, n, bias, estimate_mean, beta_true)
```
The code goes through the steps

1. Create a new variable `bias` with the difference between the estimated coefficients and the true coefficients.
2. Select the variables of interest
3. Sort by term and sample size to make it easier to understand

Use similar analysis for the other questions.

## Correlated Variables

In the previous problem, the covariates were assumed to be independent.
Now, we will evaluate the properties of OLS estimates when covariates are correlated.
As before, the population is
$$
\begin{aligned}[t]
Y_i &= 0 + 1 \cdot x_{1,i} + 1 \cdot x_{2,i} + 1 \cdot x_{3,i} + \epsilon_i \\
\epsilon_i &\sim N(0, \sigma^2) \\
\sigma &= 1.7
\end{aligned}
$$
In this problem keep $\mu_X = (0, 0, 0)$ and $s_X = (1, 1, 1)$, but $R_X$ will differ between simulations to allow for different levels of correlation between $x_1$ and $x_2$.
The covariate $x_3$ is independent of the other covariates, $\cor(x_1, x_3) = \cor(x_2, x_3) = 0$.
Thus, the correlation matrix for $X$ in these simulations is the following, where $\rho_{1,2}$ will vary:
$$
R_X =
\begin{bmatrix}
1 & \rho_{1,2} & 0 \\
\rho_{1,2} & 1 & 0 \\
0 & 0 & 1 
\end{bmatrix}
$$

Use $n = 1024$ for all simulations.

Simulate using `sim_lin_normal` with the following levels of correlation between $x_1$ and $x_2$ ($\rho_{1,2}$): 0, 0.5, 0.95, -0.5, -0.95.
Based on the results of those simulations, how does $\cor(x_1, x_2)$ affect the following?

- The bias of each $\hat{\beta}_j$?
- The variance of each $\hat{\beta}_j$?
- The bias of the standard error of each $\hat{\beta}_j$?
- The bias of the robust standard error of each $\hat{\beta}_j$?

Remember to consider the effects of correlation on *all* the estimates: $\hat{\beta}_1$, $\hat{\beta}_2$, and $\hat{\beta}_3$.

## Collinearity

Use the same settings in the previous question, but set $\cor(x_1, x_2) = 1$.
What happens?
Why?

## P-values, Type I and II errors

In this problem we will explore how $p$-values and Type I and II errors vary between samples.
Use this population model for the simulations:
$$
\begin{aligned}[t]
Y_i &= 0 + 0 \cdot x_{1,i} + 0.1 \cdot x_{2,i} + 0.5 \cdot x_{3,i} + 1 \cdot x_{4,i} + \epsilon_i \\
\epsilon_i &\sim N(0, 1.9^2)
\end{aligned}
$$
The covariates, $X$, are independent, with means, $\mu_X = (0, 0, 0)$, and standard deviations, $s_X = (1, 1, 1)$.
Use $n = 128$.
<!-- \sigma^2 = 1.9 is R^2 of approx 0.25 -->

- For each parameter, plot the distribution of $p$-values across iterations.
  Describe the densities.
- For each parameter, calculate the proportion of $p$-values less than 0.05 for each parameter.
  Suppose a null hypothesis of $H_0: \beta_i = 0$ and an alternative hypothesis of $H_a: \beta_i \neq 0$.
  For which parameters is $H_0$ true, for which parameters is $H_a$ true?
  What is the probability of a Type I or Type II error for each parameter.
- For each parameter, plot the distribution of its estimates *conditional* on it being significant at $p < 0.05$.
  How does the sampling distribution conditional on statistical significance relate to the unconditional sampling distribution?
- Repeat each of those analyses using a smaller sample size, $n = 32$, and a larger sample size $n = 1024$.
  What changes, if anything?

## Omitted Variable Bias

The population is
$$
\begin{aligned}[t]
Y_i &= 0 + 1 \cdot x_{1,i} + 1 \cdot x_{2,i} + 1 \cdot x_{3,i} + \epsilon_i \\
\epsilon_i &\sim N(0, \sigma^2) \\
\sigma &= 1.7
\end{aligned}
$$

In all simulations, $(x_1, x_2)$ and $(x_2, x_3)$ are uncorrelated.
The correlation between $x_1$ and $x_3$ will vary between simulations.
In other words, the correlation matrix for the $x$ variables is
$$
R =
\begin{bmatrix}
1 & 0 & \rho_{1,3} \\
0 & 1 & 0 \\
\rho_{1,3} & 0 & 1 
\end{bmatrix}
$$

In all simulations, the sample regression will only include $x_1$ and $x_2$:
$$
y_i = \hat\beta_0 + \hat\beta_1 x_{1,i} + \hat\beta_2 x_{2,i} + \hat\epsilon_i
$$
Use $n = 1024$ for all simulations.

You can perform this simulations of this scenario using the function, `sim_lin_norm_omitted`, which samples from a linear, normal model with homoskedastic errors, but allows you to run a misspecified regression on the samples by omitting variables.
The arguments to this function are the same as `sim_lin_norm` except for

`omit`

:    A vector of integers of the columns to omit from X when estimating the sample regression. Set `omit = 3` to omit the 3rd covariate.

```{r sim_lin_norm_omitted, echo = FALSE}
```

To run a simulation for a single value of $\rho_{1,3}$:
```{r, results='hide'}
n <- 1024
mu_X <- c(0, 0, 0)
s_X <- c(1, 1, 1)
rho <- 0
R_X <- matrix(c(1, 0, rho,
                0, 1, 0,
                rho, 0, 1), byrow = TRUE, nrow = 3)
beta <- c(0, 1, 1, 1)
sigma <- 1.7
sim_lin_norm_omitted(iter = iter, n = n, mu_X = mu_X,
                     s_X = s_X, R_X = R_X, beta = beta,
                     sigma = sigma, omit  = 3)
```

Warning: when using `summarize_sim` remember that the parameters of the population are not equal the parameters that are estimated.

Consider the following values of the correlation between $x_1$ and $x_3$ ($\rho_{1,3}$): 0, 0.1, 0.7, -0.7, 0.99, -0.99.
How does the correlation of variables affect:

- The bias of each $\hat{\beta}_j$?
- The variance of each $\hat{\beta}_j$?
- The bias of the standard error of each $\hat{\beta}_j$?
- The bias of the robust standard error of each $\hat{\beta}_j$?

## Heteroskedasticity

In this problem, we will explore how heteroskedasticity affects OLS estimates.
In these simulations, we will use a population model with heteroskedasticity
$$
\begin{aligned}[t]
Y_i &= 0 + 1 \cdot x_1 + 1 \cdot x_2 \epsilon_i \\
\epsilon_i &\sim N(0, \sigma^2_i) \\
\log \sigma_i^2 &= \gamma_0 + \gamma_1 x_1 + \gamma_2 \cdot x_2
\end{aligned}
$$

```{r echo = FALSE, eval = FALSE, results = 'hide'}
# How to find the values of heterosked
# Solve for gamma1
f <- function(ratio, gamma0, x1, x2, interval = c(0, 10)) {
  g <- function(gamma1) {
    s1 <- exp(gamma0 + gamma1 * x1)
    s2 <- exp(gamma0 + gamma1 * x2)
    ((s2 / s1) - ratio)^2
  }
  optimize(g, interval = interval)$minimum
}
f(1, 0.8, -2, 2)
f(3, 0.8, -2, 2)
f(20, 0.8, -2, 2)
```

These simulations will set $\gamma_0 = 0.8$, $\gamma_2  = 0$, and vary the level of heteroskedasticity by varying the level of $\gamma_1$.
Thus, in these simulations, $\sigma^2$ will vary with $x_1$, but not with respect to $x_2$. 
Consider the following values of $\gamma_1$:

- $\gamma_1 = 0$: no heteroskedasticity
- $\gamma_1 = 0.3$: low heteroskedasticity, $\max(\sigma^2_i) / \min(\sigma^2_i) \approx 2.7$.
- $\gamma_1 = 0.7$: high heteroskedasticity, $\max(\sigma^2_i) / \min(\sigma^2_i) \approx 24$.

The function `sim_lin_norm_hereosked` will run simulations using those specifications of the population ans ample.
The arguments for it are the same as `sim_lin_norm` except that instead of the argument `sigma` for the standard deviation of the errors, it has the argument `gamma`:

`gamma`

:    A vector of coefficients for $\log \sigma^2 = \gamma_0 + \gamma_1 x_1 + \dots + \gamma_k x_k$. Like `beta` it should have length `k + 1`.

```{r sim_lin_norm_heterosked, echo = FALSE}
```
```{r sample_lin_norm_heterosked, echo = FALSE}
```

First, set the values for this simulation:
```{r}
n <- 1024
mu_X <- c(0, 0)
s_X <- c(1, 1)
R_X <- diag(2)
beta <- c(0, 1, 1)
gamma0 <- 1.5
gamma2 <- 0
```
Create a vector when $\gamma_1 = 0$,
```{r}
gamma1_values <- c(0, 0.2, 0.7)
```

To give a better sense of what heteroskedasticity at various values of $\gamma_1$ correspond to, the following plot displays a scatter plot and a regression line for different values of $\gamma_1$:
```{r echo = FALSE}
hsk_samples <- data.frame(gamma1 = gamma1_values) %>%
  group_by(gamma1) %>%
  do({
    sample_lin_norm_heterosked(1, n = 512, mu_X = 0,
                               s_X = 1, R_X = matrix(1),
                               beta = c(0, 1),
                               gamma = c(gamma0, .$gamma1))
  })

ggplot(hsk_samples, aes(x = V1, y = y)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ gamma1, ncol = 2) +
  ylab("X") + 
  theme_minimal()
```

For given values, you can run `sim_lin_norm_heterosked`:
```{r results='hide'}
gamma1 <- gamma1_values[1]
gamma <- c(gamma0, gamma1, gamma2)
sim_lin_norm_heterosked(iter = iter, n = n, mu_X = mu_X,
                        s_X = s_X, R_X = R_X, beta = beta, gamma = gamma)
```
For these simulations, you will need to loop over the various values of $\gamma_1$ stored in `gamma1_values`.

Use `sim_lin_norm_heterosked` to evaluate how the level of heteroskedasticity affects:

- the bias of each $\hat{\beta}_j$?
- the variance of each $\hat{\beta}_j$?
- the bias of the standard error of each $\hat{\beta}_j$?
- the bias of the robust standard error of each $\hat{\beta}_j$?

Remember to consider how heteroskedasticity in $x_1$ affects the estimates of heteroskedasticity in $x_2$.

## Truncated Dependent Variable (optional)


This problem considers what happens when there is a truncated dependent variable.
This is also called sampling on the dependent variable, which is a research design problem not unknown to political science research.[^samplingdv]

The population is a linear normal model with homoskedastic errors.
$$
\begin{aligned}[t]
Y_1 &= \beta_0 + \beta_1 x_{1,i} + \dots + \beta_k x_{k,i} + \epsilon_i \\
\epsilon_i &\sim N(0, \sigma^2)
\end{aligned}
$$
However, in each sample, all $y_i$ which are less than a quantile $q$ are dropped before the regression is estimated.
$$
\begin{aligned}[t]
y_i = \beta_0 + \hat\beta_1 x_{1,i} + \dots + \hat\beta_k x_{k,i} + \hat\epsilon \\ \text{if $y_i \geq \quantile(y, q)$}
\end{aligned}
$$
where $\quantile(y, q)$ is the $q$th quantile of $y$.
For example, if $q = 0.5$, all $y_i$ that are less than the median of $y$ (the bottom 50%) are dropped.

The function `sim_lin_norm_truncated`, is similar to the `sim_lin_norm` function except for the argument

`truncation`

:    The quantile of truncation. All sampled $y_i$ with values less than that quantile are dropped before the regression is run.
     The default value `truncation = 0.5` means all values of $y$ less than the median are dropped before running the regression.

```{r sim_lin_norm_truncated, echo = FALSE}
```

<!--
For this problem, use only a single covariate, with $\mu_X = 0$,$s_X = 1$, and $R_X = 1$.
Let $\beta = (0, 1)$ and $\sigma = 1$.
Estimate the regression only on those $y_i$ greater than the median of $y$.
-->

Before running simulations, draw a single sample of a linear normal model with homoskedastic errors.
To do this, you should be able to adapt the code from `sim_lin_normal_truncated`.
Create a scatter plot with the OLS line for all $y$, and a another plot with only those $y$ less than the median of $y$.
How does the OLS line estimated on the truncated data differ from the one estimated on the full data.

Run several simulations with `sim_lin_normal_truncated` and vary the sample size.
How does the sample size affect the following:

- The bias of each $\hat{\beta}_j$?
- The variance of each $\hat{\beta}_j$?

In particular, if we gather more data but $y$ is truncated, does it decrease the bias in $\hat{\beta}$?

[^randomx]: Although the statistical theory of OLS works (thankfully) for random $X$,
    as long as certain conditions are met. See Fox (2nd ed.), Ch 9.6.

[^samplingdv]: See Ashworth, Scott, Joshua D. Clinton, Adam Meirowitz, and Kristopher W. Ramsay. 2008. ``Design, Inference, and the Strategic Logic of Suicide Terrorism.'' *American Political Science Review() 102(02): 269–73. <http://journals.cambridge.org/article_S0003055408080167>

